---
title: "Choosing an interval"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an interval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`reintervals` computes four interval types for a linear mixed model, and picks
the computational engine automatically from the fitted model. This vignette is
the map: given your model, which intervals are available, which method you get
by default, and what to watch for.

```{r setup, message = FALSE}
library(reintervals)
library(lme4)
```

## The four intervals

| Verb | Answers |
|---|---|
| `ci_lmm()` | Confidence interval for the **mean**. |
| `pi_lmm()` | Prediction interval for a **single future observation**. |
| `ti_lmm()` | Tolerance interval covering a **proportion `P` of the population**. |
| `new_group_mean_lmm()` | Interval for the **mean of one new, as-yet-unobserved group** (CInew / new-batch-mean). |

## The four engines

| Engine (`method`) | Used for | Reference |
|---|---|---|
| `ems-mls` | Random-intercept designs and the balanced fixed-slope design (expected-mean-square / observed-information MLS). | Francq, Lin & Hoyer (2019) |
| `anova-mls` | Single-observation / unbalanced fixed-slope (stability) tolerance intervals (unweighted lot-mean ANOVA). | Montes, Burdick & Leblond (2019) |
| `reml-mls` | Random-slope closed-form intervals. | Graybill–Wang MLS two-component bound (Graybill & Wang 1980); generalized Satterthwaite df |
| `gpq` | Random-slope batch-mean target, and singular random-slope tolerance intervals (generalized pivotal quantity, Monte-Carlo). | Oliva-Aviles & Hauser (2025) |

The engine is chosen automatically from the model's structure; `method =` lets
you pin it (within what is valid — see below).

## Which interval for which model

Read down to your model structure. `ems`, `anova`, `reml`, `gpq` are the engine
tags; superscript letters are explained below.

| Model / situation | CI | PI | TI | CI_new |
|---|---|---|---|---|
| Random intercept — one-way (balanced or unbalanced) | `ems` | `ems` | `ems` | `ems` |
| Random intercept — nested (balanced or unbalanced) | `ems` | `ems` | `ems` | `ems` |
| Random intercept — crossed **with** interaction (balanced/unbalanced; >2 VCs) | `ems` | `ems` | `ems` | `ems` |
| Random intercept — crossed **no** interaction | `ems` | `ems` | **error** ^a^ | `ems` |
| Random intercept — ≥4 grouping factors (off-catalog) | `ems` | `ems` | **error** ^a^ | `ems` |
| Random intercept + fixed slope — clean balanced | `ems` | `ems` | `ems` | `ems` |
| Random intercept + fixed slope — single-obs / staggered / unbalanced | `ems` ^b^ | `ems` ^b^ | **`anova`** ^c^ | `ems` ^b^ |
| Random slope — non-singular, future-observation target | `reml` | `reml` | `reml` | `reml` |
| Random slope — non-singular, batch-mean target (`over = "group_mean"`) | — ^d^ | — ^d^ | **`gpq`** | — ^d^ |
| Random slope — singular fit | `reml` ^e^ | `reml` ^e^ | **`gpq`** ^e^ | `reml` ^e^ |
| Random slope + ≥2 fixed effects | **error** ^f^ | **error** ^f^ | **error** ^f^ | **error** ^f^ |

^a^ These designs have no closed-form variance decomposition, so `ti_lmm()`
signals a clear error. CI, PI, and CI_new remain available.

^b^ On single-observation / unbalanced fixed-slope data the EMS coverage
intervals can be anti-conservative (too narrow); `ci_lmm()`, `pi_lmm()`, and
`new_group_mean_lmm()` emit a warning. The interval is returned unchanged.

^c^ `ti_lmm()` re-routes to the Montes ANOVA closed form (`anova-mls`), which is
validated for this structure.

^d^ `over = "group_mean"` (batch-mean) is a `ti_lmm()` argument only, and the
batch-mean *tolerance* target is random-slope-only in this version. For other
designs, `new_group_mean_lmm()` gives an interval for a new batch's mean — it is
available for random-intercept, fixed-slope, and random-slope models. The
coverage verbs carry no batch-mean target.

^e^ For a singular random-slope fit, the tolerance interval switches to the GPQ
engine (the closed form's batch-mean/singular behavior is not validated); CI, PI,
and CI_new stay on the closed form, flagged `singular`.

^f^ Random-slope support is limited to a single continuous covariate (`~ 1 + x`)
in this version — time in the stability use case, but any continuous predictor works.

## Choosing the method by hand

By default `method = "auto"` selects the one engine valid for your model. For
most models there is no alternative — requesting a different engine errors. Only
two cases offer a genuine choice, both for tolerance intervals:

- **Single-observation fixed-slope TI:** default `"anova-mls"` (Montes, validated
  for this structure); `"ems-mls"` forces the EMS engine with no re-route — but
  that's the path flagged as anti-conservative here, so it's mainly for comparison.
- **Random-slope TI (future-observation target):** default `"reml-mls"` (closed
  form) on a non-singular fit; `"gpq"` (Monte-Carlo) is also valid and becomes the
  default automatically when the fit is singular. For the batch-mean target, only
  `"gpq"` is available.

## Worked examples


A random-intercept + fixed-slope model — every verb uses `ems-mls`:

```{r ex-ems}
fm <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)
ci_lmm(fm, newdata = data.frame(Days = 5))
ti_lmm(fm, newdata = data.frame(Days = 5), P = 0.99)
```

A random-slope model — the verbs use the `reml-mls` closed form:

```{r ex-slope}
fs <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)
ci_lmm(fs, newdata = data.frame(Days = 5))
```

The batch-mean tolerance interval on the same slope model uses GPQ (Monte-Carlo
set a seed for reproducibility):

```{r ex-gpq}
ti_lmm(fs, newdata = data.frame(Days = 5), P = 0.99,
       over = "group_mean",
       control = lmm_interval_control(seed = 1, M = 20000))
```

## Limitations in this version

- **One-sided intervals** (`sides = "lower"` / `"upper"`) are available only for
  the Montes single-observation tolerance interval and the random-slope engines.
  The EMS engine (random-intercept and balanced fixed-slope) is two-sided only.
- **Tolerance intervals** require a closed-form variance decomposition; for
  crossed-no-interaction and ≥4-factor designs, `ti_lmm()` errors (CI/PI/CI_new
  remain available).
- **Batch-mean tolerance intervals** (`over = "group_mean"`) are currently
  available only for random-slope models; a random-intercept batch-mean tolerance
  interval is planned for a future release.
- **Random-slope** models are supported for a single continuous covariate (time
  in the stability design, but any continuous predictor works).

## References

- Francq, B. G., Lin, D., & Hoyer, W. (2019). Confidence, prediction, and
  tolerance in linear mixed models. *Statistics in Medicine*, 38(30), 5603–5622.
  <doi:10.1002/sim.8386>
- Montes, R. O., Burdick, R. K., & Leblond, D. J. (2019). Simple approach to
  calculate random effects model tolerance intervals to set release and
  shelf-life specification limits of pharmaceutical products. *PDA Journal of
  Pharmaceutical Science and Technology*. <doi:10.5731/pdajpst.2018.008839>
- Graybill, F. A., & Wang, C.-M. (1980). Confidence intervals on nonnegative
  linear combinations of variances. *Journal of the American Statistical
  Association*, 75(372), 869–873.
- Oliva-Aviles, C., & Hauser, P. (2025). *Technometrics*, 67(2), 193–202.
  <doi:10.1080/00401706.2024.2407324>
