---
title: "Getting started with reintervals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with reintervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
library(reintervals)
```

`reintervals` computes four kinds of interval for a linear mixed model fitted
with `lme4::lmer()`, through one uniform interface, and returns them in one tidy
`lmm_interval` data frame.

## 1. The four interval types --- and when to use each

| Verb | Answers | Use when you want... |
|---|---|---|
| `ci_lmm()` | confidence interval for the **mean** | uncertainty in the average response at a point |
| `pi_lmm()` | prediction interval for **one future observation** | where a single new measurement will fall |
| `ti_lmm()` | tolerance interval covering a proportion `P` of the **population** | a content/coverage statement (e.g. "99% of units, 95% confident") |
| `new_group_mean_lmm()` | the **CInew / new-batch-mean** interval | where the mean of one new, unobserved group/batch will fall |

`lmm_predict(model, interval = ...)` is a `predict()`-style wrapper over the
four. The engine that computes each interval is chosen automatically from the
fitted model (you can override with `method = `).

## 2. The four verbs on a random-intercept model

We use `sleepstudy` (reaction time by `Days`, random intercept per `Subject`):

```{r ri-fit}
fm <- lme4::lmer(Reaction ~ Days + (1 | Subject), data = lme4::sleepstudy)

ci_lmm(fm, newdata = data.frame(Days = 5))
pi_lmm(fm, newdata = data.frame(Days = 5))
ti_lmm(fm, newdata = data.frame(Days = 5), P = 0.99)
new_group_mean_lmm(fm, newdata = data.frame(Days = 5))
```

### Reading the output

Every verb returns the same `lmm_interval` columns:

* the **eval-point column(s)**, named by the actual predictor (`Days` here);
* `type` (`ci`/`pi`/`ti`/`new_group_mean`), `estimate`, `lower`, `upper`;
* `level` (the confidence, `1 - alpha`) and `P` (content proportion --- a value
  for `ti`, `NA` for the coverage intervals); `sides`;
* `method` (the engine that ran) and `design` (the route-independent model
  class);
* the per-row flags `singular`, `boundary`, `vg_floored`.

Engine-specific extras (degrees of freedom, GPQ/ANOVA intermediates) live in an
attribute, and the random-intercept topology in another:

```{r ri-attrs}
ci <- ci_lmm(fm, newdata = data.frame(Days = 5))
str(attr(ci, "diagnostics"))
attr(ci, "grouping")
```

## 3. Stability worked examples that double as validation

The next two examples are **not illustrations** --- they are the package
reproducing its published validation oracles, live.

### Montes-Burdick-Leblond (2019) Appendix A

`mbl_appendix_a` is the committed Appendix A dataset (11 lots, mixing
multi-timepoint stability lots with release-only single-observation lots). The
published two-sided 99%-content / 95%-confidence tolerance interval at month 12
is **(57.25, 75.41)**:

```{r mbl}
head(mbl_appendix_a)
fm_mbl <- lme4::lmer(Y ~ Month + (1 | Lot), data = mbl_appendix_a)
ti_lmm(fm_mbl, newdata = data.frame(Month = 12), P = 0.99)
```

The dispatcher routes this single-observation design to the unweighted lot-mean
ANOVA engine (`method = "anova-mls"`), which reproduces the published number.

### Oliva-Aviles & Hauser (2025) GPQ

`oliva_hauser_dgp` is the published Section-5 data-generating specification. The
content variance the GPQ method targets at `t0 = 20` has the closed form
**`tau^2 = 8.439`**:

```{r gpq-closed}
oliva_hauser_dgp$tau2_at_t0
```

The GPQ engine is Monte Carlo: its `tau^2` estimate (`tau2_med`) is noisy on a
single dataset but recovers the closed-form value **in expectation**. Simulating
20 datasets from the published DGP and averaging:

```{r gpq-recovery}
d <- oliva_hauser_dgp
sim_one <- function(seed, K = 40) {
  set.seed(seed)
  Lc <- chol(matrix(c(d$s2_0, d$s01, d$s01, d$s2_1), 2, 2))
  do.call(rbind, lapply(seq_len(K), function(k) {
    a <- as.numeric(crossprod(Lc, rnorm(2)))                 # batch random effects
    data.frame(batch = k, time = d$times,
               y = (d$beta0 + a[1]) + (d$beta1 + a[2]) * d$times +
                   rnorm(length(d$times), 0, sqrt(d$s2e)))
  }))
}
tau2_med <- vapply(1:20, function(s) {
  x <- sim_one(s)
  g <- ti_gpq_raw(x$y, x$time, x$batch, t0 = 20, P = 0.90, level = 0.95,
                  c_resid = 0, control = lmm_interval_control(seed = s, M = 2000))
  attr(g, "diagnostics")$tau2_med
}, numeric(1))

c(closed_form = d$tau2_at_t0, mean_gpq_tau2 = mean(tau2_med))
```

The mean GPQ `tau^2` recovers the published `8.439` (the package's test suite
checks both the exact closed form and this expectation on CI).

## 4. The engine / scope map

The engine is chosen from the fitted model's structure (full mapping in
`?reintervals-models`):

| Model class | Engine (`method`) |
|---|---|
| random intercept (any fixed effects) | `ems-mls` |
| random intercept + fixed slope, clean balanced | `ems-mls` |
| random intercept + fixed slope, single-observation / unbalanced | `anova-mls` |
| random slope, future-observation | `reml-mls` (`gpq` if singular) |
| random slope, batch-mean (`over = "group_mean"`) | `gpq` |

**v1 scope:** general fixed effects for random-intercept models; a **single time
covariate** for random-slope models (the stability design).

### A warning you should understand

On a fixed-slope model fitted to single-observation / unbalanced lots, the EMS
confidence/prediction/new-group-mean intervals can be **anti-conservative (too
narrow)**, because REML precision-weighting down-weights the single-observation
lots. The coverage verbs **warn** on this structure (the interval is still
returned). Here the warning fires on the MBL data:

```{r c1-warning}
ci_lmm(fm_mbl, newdata = data.frame(Month = 12))
```

The *tolerance* interval is the one re-routed to the validated Montes ANOVA
closed form on this structure (Section 3); a validated closed form for the
CI/PI/new-group-mean case is planned for v2.

## 5. Limitations

* **The `vg_floored` flag is a defensive guard you should not normally see.**
  A valid `lmer` fit has a positive-semidefinite random-effect covariance, so the
  between-group variance `V_G(t0) = z0' Sigma z0` is non-negative at every `t0`.
  The floor-and-flag only triggers if a (numerically degenerate) fit drives it
  below zero; in normal use it stays `FALSE`/`NA`. Do not be alarmed by a guard
  you will not hit.

* **`V_G` is quadratic in the time covariate.** Evaluating a slope interval at a
  `t0` outside the observed time range extrapolates that quadratic; do it
  knowingly. There is no guard or warning for extrapolation.

## 6. Where to go next

* `?reintervals` --- package overview and the engine/scope map.
* `?reintervals-models` --- the model-class to engine mapping and design tags.
* `?ci_lmm` --- the full `lmm_interval` output contract (shared by all verbs).
* `NEWS.md` --- the change log and the v2 roadmap (multi-covariate random-slope
  intervals).
